FEA-Comp: an Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-03-24 paper/bare_conf.tex
Modifications + related work

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations}

%\author{\IEEEauthorblockN{Fabio Luporini}
%\IEEEauthorblockA{Department of Computing\\
%Imperial College London\\
%London, UK\\
%Email: f.luporini12@imperial.ac.uk}
%\and
%\IEEEauthorblockN{Florian Rathgeber}
%\IEEEauthorblockA{Department of Computing\\
%Imperial College London\\
%London, UK\\
%Email: f.rathgeber10@imperial.ac.uk}
%\and
%\IEEEauthorblockN{Ana Lucia Varbanescu}
%\IEEEauthorblockA{Department of Computing\\
%Imperial College London\\
%London, UK\\
%Email: a.l.varbanescu@uva.nl}
%\and
%\IEEEauthorblockN{Gheorghe-Teodor Bercea}
%\IEEEauthorblockA{Department of Computing\\
%Imperial College London\\
%London, UK\\
%Email: gb308@doc.ic.ac.uk}
%\and
%\IEEEauthorblockN{Paul H. J. Kelly}
%\IEEEauthorblockA{Department of Computing\\
%Imperial College London\\
%London, UK\\
%Email: p.kelly@imperial.ac.uk}
%\and
%\IEEEauthorblockN{David Ham}
%\IEEEauthorblockA{Department of Computing\\
%Imperial College London\\
%London, UK\\
%Email: d.ham@imperial.ac.uk}}

\author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
Florian Rathgeber\IEEEauthorrefmark{1},
Ana Lucia Varbanescu\IEEEauthorrefmark{2}, 
Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
Paul H. J. Kelly\IEEEauthorrefmark{1},
and
David Ham\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
\IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham@imperial.ac.uk\}@imperial.ac.uk,\\
\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl
}
}

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $icc$ and $gcc$ have. First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension. This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

Table~\ref{table:perf-results}

\newcommand{\massmatrixtwodresultsnorms}{
\begin{tabularx}{1.0\textwidth}{cccc|cc|cc|}
\cline{5-8}
& & & & \multicolumn{2}{c|}{$split$} & \multicolumn{2}{c|}{$op$-$vect$} \\
\cline{2-8}
& \multicolumn{1}{|c|}{original} & \multicolumn{1}{|c|}{$licm$} & $licm$+$ap$ & Best & Cost Model & Best & Cost Model\\
\cline{1-8}
\multicolumn{1}{|c|}{Helmholtz$\_$tri$\_$p1} & & \multicolumn{1}{|c|}{$5.49e{-}17$} & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\cline{1-8}
\end{tabularx}
}
\begin{figure*}
\massmatrixtwodresultsnorms
\end{figure*}

@@

\section{Related Work}
\label{sec:related-work}
%Show other work of people working on optimizations for computational science kernels. Inter-kernel vectorisation paper from Istvan. Previous work in FFC. Spencer et-al + Shin et. al. attempt to optimize computations that could benefit from using BLAS, but in practice they don't, due to the very small dgemm operations employed. Saday's model-driven SIMD code vectorisation for the tensor contraction engine.

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{FEA-Comp: an Optimizing Compiler\\for Solving Partial Differential Equations}

\author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
Florian Rathgeber\IEEEauthorrefmark{1},
Ana Lucia Varbanescu\IEEEauthorrefmark{2}, 
Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
David Ham\IEEEauthorrefmark{1},
and
Paul H. J. Kelly\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
\IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham@imperial.ac.uk\}@imperial.ac.uk,\\
\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl
}
}

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $icc$ and $gcc$ have. First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension. This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

Tables~\ref{table:perf-results-licm-helmholtz},~\ref{table:perf-results-licm-advdiff}, and~\ref{table:perf-results-licm-burgers} illustrate the speed-ups obtained on the Sandy Bridge and the Phi machines when $licm$, data alignment, and padding are used, over non-transformed code. Inspection of assembly code generated by $icc$ confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant with respect to outer loops are hoisted and, interestingly, not vectorized. This motivates the usually significant gain. Padding and data alignment enhance, in general, the quality of SIMD auto-vectorization. Sometimes the value of $licm$-$ap$ is similar to that of $licm$ because the stiffness matrix size is, without padding, already a multiple of the vector length, and data is automatically aligned. Occasionally $licm$-$ap$ is slower than $licm$ (e.g. in Burgers 3D p3). This is due to the large number of aligned temporaries introduced by $licm$, which probably lead to cache associativity conflicts.

\newcommand{\massmatrixtwodresultsnorms}{
\begin{tabularx}{1.0\textwidth}{p{3.5cm}|p{1.98cm}p{1.98cm}p{1.98cm}|p{1.98cm}p{1.98cm}p{1.98cm}|}
\cline{2-7}
& \multicolumn{3}{c|}{\texttt{AVX}} & \multicolumn{3}{c|}{\texttt{PHI}} \\
\cline{2-7}
& \texttt{original} & \texttt{licm} & \texttt{licm+ap} & \texttt{original} & \texttt{licm} & \texttt{licm+ap}\\
\cline{1-7}
\multicolumn{1}{|c|}{\texttt{triangle p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{triangle p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{triangle p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{triangle p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\cline{1-7}
\multicolumn{1}{|c|}{\texttt{tetrahedron p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{tetrahedron p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{tetrahedron p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{tetrahedron p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\cline{1-7}
\multicolumn{1}{|c|}{\texttt{prism p1}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{prism p2}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{prism p3}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\multicolumn{1}{|c|}{\texttt{prism p4}} & $5.49e{-}17$ & $4.95e{-}17$ & $1.44e{-}17$ & $1.34e{-}17$ & $1.56e{-}17$ & $\mathbf{4.43e{-}5}$ \\
\cline{1-7}

\end{tabularx}
}

\begin{table*}[t]\normalsize
\massmatrixtwodresultsnorms
\label{table:perf-results-licm-helmholtz}
\caption{Impact of generalized loop-invariant code motion ($licm$ column) on the Helmholtz problem, for three kind of elements belonging to the Conjugate Gradient family (triangle, tetrahedron, prism) and for the range of polynomial orders $p$ $\in$ $\lbrace$1, 4$\rbrace$. The column $licm$-$ap$ illustrates the combination of $licm$ with data alignment and padding.}
\end{table*}

@@

\section{Related Work}
\label{sec:related-work}
The finite element method is used in the most disparate contexts to approximate solutions of PDEs. Well-known frameworks and applications include~\cite{nek5000-web-page},~\cite{Fenics},~\cite{fluidity_manual_v4}, and of course Firedrake; this is not an exhaustive list, though. Different techniques and algorithms are used depending on the level of accuracy that needs to be reached. When low-order polynomials are used, standard numerical integration is usually employed. The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches. For instance, methods based on tensor contraction~\cite{FFC-TC} and symbolic manipulation~\cite{Francis} have been implemented, although numerical integration remains the optimal choice for a wide class of problems~\cite{quadrature1}.

Optimization of local assembly using numerical integration for CPU platforms has been tackled in~\cite{quadrature1}. However, to the best of our knowledge, ours is the first work targeting low-level compiler optimizations. In~\cite{Markall20101815}, and more recently in~\cite{petsc-integration-gpu}, the problem has been studied for GPUs. In~\cite{assembly-opencl}, variants of the standard numerical integration algorithm are specialized for the PowerXCell processor and evaluated; it lacks, however, an exhaustive study from the compiler viewpoint as we did, and none of the optimizations presented in Section~\ref{sec:code-transf} is discussed.

Domain-specific languages and automated code generation have been adopted with success in different fields. Spiral~\cite{Pueschel:05} automates generation of highly-optimized platform-specific digital signal processing numerical algorithms. Similar ideas are adopted in~\cite{Spampinato:14}, which proposes DSLs and compilers for dense linear algebra kernels. OP2~\cite{op2-main}, and its python implementation~\cite{pyop2isc}, which is also used by Firedrake, aim at achieving performance portability for scientific computations based on unstructured meshes.

- finding the best problem implementation: cost model based vs auto-tuning vs hybrid

%(i.e. given a mesh, those that lead to ``less accurate'' solutions than when using high-order polynomials, at a lower computational cost)

%Show other work of people working on optimizations for computational science kernels. Inter-kernel vectorisation paper from Istvan. Previous work in FFC. Spencer et-al + Shin et. al. attempt to optimize computations that could benefit from using BLAS, but in practice they don't, due to the very small dgemm operations employed. Saday's model-driven SIMD code vectorisation for the tensor contraction engine.
